---
title: "2016 Claims-Based Frailty Index Descriptive Analysis"
author: "Nemo Zhou"
date: "2026-06-17"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

<!--
Project: Frailty_Komoto 2016 CFI descriptive visualization
Author: Nemo Zhou
Date started: 2026-06-17
Date last updated: 2026-06-18

Purpose:
This R Markdown report visualizes aggregate 2016 Claims-Based Frailty Index
(CFI) descriptive outputs created by:
Code/4.1_extract_2016_cfi_descriptive_outputs.R

It does not connect to Redshift. It reads aggregate CSV files from the
project-root Outputs folder and renders a frailty-group Table 1, CFI summary
table, an overall histogram, and box plots.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)

library(ggplot2)
library(knitr)
```

```{r helpers}
output_dir_candidates <- c("Outputs", "../Outputs")
output_dir <- output_dir_candidates[
  dir.exists(output_dir_candidates)
][1]

if (is.na(output_dir)) {
  stop(
    "Could not find the project-root Outputs folder. ",
    "Knit this report from the project root or the Code folder."
  )
}

read_required_csv <- function(file_name) {
  path <- file.path(output_dir, file_name)
  if (!file.exists(path)) {
    stop(
      "Missing required file: ",
      path,
      ". Run Code/4.1_extract_2016_cfi_descriptive_outputs.R first."
    )
  }

  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

metadata <- read_required_csv("4.1_cfi_2016_metadata.csv")

metadata_value <- function(field) {
  value <- metadata$value[metadata$field == field]
  if (length(value) == 0L) {
    return(NA_character_)
  }
  value[[1]]
}

comma_label <- function(x) {
  format(
    round(x),
    big.mark = ",",
    scientific = FALSE,
    trim = TRUE
  )
}

cfi_label <- function(x) {
  sprintf("%.2f", x)
}

extract_count <- function(x) {
  cleaned <- gsub(",", "", x)
  cleaned <- sub(" .*", "", cleaned)
  suppressWarnings(as.numeric(cleaned))
}

extract_percent <- function(x) {
  ifelse(
    grepl("\\(", x),
    sub(".*\\(([^)]+)\\).*", "\\1", x),
    x
  )
}

format_percent_one_decimal <- function(x) {
  percent_value <- suppressWarnings(
    as.numeric(gsub("%", "", as.character(x)))
  )

  ifelse(
    is.na(percent_value),
    as.character(x),
    paste0(sprintf("%.1f", percent_value), "%")
  )
}

format_percent_columns <- function(data) {
  percent_columns <- grepl("percent|percentage", names(data), ignore.case = TRUE)

  for (column in names(data)[percent_columns]) {
    data[[column]] <- format_percent_one_decimal(data[[column]])
  }

  data
}

format_numeric_columns <- function(data, columns, digits) {
  for (column in intersect(columns, names(data))) {
    numeric_value <- suppressWarnings(as.numeric(data[[column]]))
    data[[column]] <- ifelse(
      is.na(numeric_value),
      as.character(data[[column]]),
      sprintf(paste0("%.", digits, "f"), numeric_value)
    )
  }

  data
}

format_table_one_percent_only <- function(data) {
  frailty_columns <- setdiff(names(data), c("variable", "level"))

  overall_row <- data[
    data$variable == "Overall" & data$level == "Overall",
    ,
    drop = FALSE
  ]

  if (nrow(overall_row) == 1L) {
    new_names <- names(data)
    for (column in frailty_columns) {
      count <- extract_count(overall_row[[column]])
      if (!is.na(count) && !grepl("\\(N=", column, fixed = FALSE)) {
        new_names[new_names == column] <- paste0(
          column,
          " (N=",
          format(
            round(count),
            big.mark = ",",
            scientific = FALSE,
            trim = TRUE
          ),
          ")"
        )
      }
    }
    names(data) <- new_names
    frailty_columns <- setdiff(names(data), c("variable", "level"))
  }

  for (column in frailty_columns) {
    data[[column]] <- format_percent_one_decimal(extract_percent(data[[column]]))
  }

  data
}
```

```{r read-data}
table_one_display <- read_required_csv(
  "4.1_cfi_2016_table1_by_frailty_group.csv"
)
table_one_display <- format_table_one_percent_only(table_one_display)

subgroup_summary <- read_required_csv("4.1_cfi_2016_subgroup_summary.csv")
if ("n_patient_years" %in% names(subgroup_summary)) {
  names(subgroup_summary)[names(subgroup_summary) == "n_patient_years"] <- "n_patient"
}
subgroup_summary <- format_percent_columns(subgroup_summary)
subgroup_summary <- format_numeric_columns(
  subgroup_summary,
  c(
    "mean_cfi",
    "minimum_cfi",
    "q1_cfi",
    "median_cfi",
    "q3_cfi",
    "maximum_cfi"
  ),
  digits = 3
)

histogram_data <- read_required_csv("4.1_cfi_2016_histogram_bins.csv")
boxplot_data <- read_required_csv("4.1_cfi_2016_boxplot_statistics.csv")
if ("n_patient_years" %in% names(boxplot_data)) {
  names(boxplot_data)[names(boxplot_data) == "n_patient_years"] <- "n_patient"
}
```

# Analysis Definitions

The three frailty strata used for Table 1 are:

- Non-frail: CFI `< 0.15`
- Prefrail: CFI `0.15` to `< 0.25`
- Frail: CFI `>= 0.25`

Analysis year: `r metadata_value("analysis_year")`

Minimum reported cell count: `r metadata_value("min_count")`

Race/ethnicity source: `r metadata_value("race_source")`

Histogram note: `r metadata_value("histogram_note")`

# Table 1 By Frailty Group

```{r table-one-display}
knitr::kable(table_one_display)
```

# CFI Summary By Subgroup

```{r subgroup-summary}
knitr::kable(subgroup_summary)
```

# Histogram Of CFI

```{r histogram-overall}
histogram_overall <- histogram_data[histogram_data$stratifier == "Overall", ]
histogram_x_min <- floor(min(histogram_overall$cfi_bin_start, na.rm = TRUE) * 100) / 100
histogram_x_max <- ceiling(max(histogram_overall$cfi_bin_start, na.rm = TRUE) * 100) / 100

ggplot(histogram_overall, aes(x = cfi_bin_start, y = n_patient_years)) +
  geom_col(width = 0.01, fill = "#3F7C85", color = "white", linewidth = 0.1) +
  scale_x_continuous(
    labels = cfi_label,
    breaks = seq(histogram_x_min, histogram_x_max, by = 0.05),
    minor_breaks = seq(histogram_x_min, histogram_x_max, by = 0.01)
  ) +
  scale_y_continuous(labels = comma_label) +
  labs(
    title = "Distribution of 2016 CFI Scores",
    x = "CFI score",
    y = "Eligible patient-years"
  ) +
  theme_minimal(base_size = 12)
```

# Box Plot Of CFI

```{r boxplot-overall}
boxplot_overall <- boxplot_data[boxplot_data$stratifier == "Overall", ]
overall_box_stats <- data.frame(
  statistic = c("Min", "Q1", "Median", "Q3", "Max"),
  value = as.numeric(boxplot_overall[, c("ymin", "lower", "middle", "upper", "ymax")])
)
overall_box_stats$label <- paste0(
  overall_box_stats$statistic,
  ": ",
  sprintf("%.3f", overall_box_stats$value)
)

ggplot(
  boxplot_overall,
  aes(x = 1, ymin = ymin, lower = lower, middle = middle,
      upper = upper, ymax = ymax)
) +
  geom_boxplot(stat = "identity", fill = "#D9A441", width = 0.45) +
  geom_point(
    data = overall_box_stats,
    aes(x = 1, y = value),
    inherit.aes = FALSE,
    size = 2
  ) +
  geom_text(
    data = overall_box_stats,
    aes(x = 1.28, y = value, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    size = 3.3
  ) +
  scale_x_continuous(
    breaks = 1,
    labels = "Overall",
    limits = c(0.65, 1.75)
  ) +
  labs(
    title = "2016 CFI Score Distribution",
    x = NULL,
    y = "CFI score"
  ) +
  theme_minimal(base_size = 12)
```

```{r boxplot-by-stratifier}
boxplot_by_stratifier <- boxplot_data[boxplot_data$stratifier != "Overall", ]

ggplot(
  boxplot_by_stratifier,
  aes(x = reorder(stratum, middle), ymin = ymin, lower = lower,
      middle = middle, upper = upper, ymax = ymax)
) +
  geom_boxplot(stat = "identity", fill = "#A44A3F", width = 0.5) +
  coord_flip() +
  facet_wrap(~ stratifier, scales = "free_y") +
  labs(
    title = "2016 CFI Score Distribution by Subgroup",
    x = NULL,
    y = "CFI score"
  ) +
  theme_minimal(base_size = 10)
```

# Source Files

This report reads these aggregate CSV files from `Outputs`:

- `4.1_cfi_2016_metadata.csv`
- `4.1_cfi_2016_table1_by_frailty_group.csv`
- `4.1_cfi_2016_subgroup_summary.csv`
- `4.1_cfi_2016_histogram_bins.csv`
- `4.1_cfi_2016_boxplot_statistics.csv`
