---
title: "Annual Healthcare Utilization"
author: "Nemo Zhou"
date: "2026-07-17"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
params:
  analysis_year: 2022
  utilization_output_dir: null
---

<!--
Project: Frailty_Komoto annual healthcare-utilization visualization
Author: Nemo Zhou
Date started: 2026-07-17
Date last updated: 2026-07-17

Purpose:
This CSV-only R Markdown report visualizes aggregate annual healthcare-
utilization outputs created by
Code/2_variable construction/6.4_describe_annual_healthcare_utilization.R.
It does not connect to Redshift. Visit prevalence is shown as the number and
percentage of eligible patient-years with at least one visit. Duration plots
use aggregate duration frequencies, so no patient-level or utilization-level
rows are loaded locally.
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 12,
  fig.height = 7
)

library(ggplot2)
library(knitr)
```

```{r helpers}
selected_analysis_year <- as.integer(params$analysis_year)
if (
  length(selected_analysis_year) != 1L ||
    is.na(selected_analysis_year) ||
    selected_analysis_year != 2022L
) {
  stop("Set params$analysis_year to 2022 for this fixed utilization workflow.")
}

if (!is.null(params$utilization_output_dir) && nzchar(params$utilization_output_dir)) {
  requested_output_dir <- params$utilization_output_dir
  output_dir_candidates <- c(
    requested_output_dir,
    file.path("..", requested_output_dir),
    file.path("..", "..", requested_output_dir),
    file.path("..", "..", "..", requested_output_dir)
  )
} else {
  output_dir_candidates <- c(
    file.path("Outputs", paste0("6.x_annual_healthcare_utilization_", selected_analysis_year)),
    file.path("..", "Outputs", paste0("6.x_annual_healthcare_utilization_", selected_analysis_year)),
    file.path("..", "..", "Outputs", paste0("6.x_annual_healthcare_utilization_", selected_analysis_year))
  )
}

output_dir <- output_dir_candidates[dir.exists(output_dir_candidates)][1]
if (is.null(output_dir) || is.na(output_dir) || !dir.exists(output_dir)) {
  stop(
    "Could not find the healthcare-utilization output directory. Run ",
    "Code/2_variable construction/6.4_describe_annual_healthcare_utilization.R first."
  )
}

read_required_csv <- function(file_name) {
  path <- file.path(output_dir, file_name)
  if (!file.exists(path)) {
    stop(
      "Missing required file: ", path,
      ". Run Code/2_variable construction/6.4_describe_annual_healthcare_utilization.R first."
    )
  }
  utils::read.csv(path, stringsAsFactors = FALSE, check.names = FALSE)
}

summary_required_columns <- c(
  "analysis_year", "stratification", "stratum_value", "utilization_category",
  "n_eligible_patient_years", "n_patients_with_visit",
  "pct_patient_years_with_visit", "n_visits", "duration_n",
  "duration_missing_n", "duration_missing_pct", "duration_mean_days",
  "duration_sd_days", "duration_min_days", "duration_p25_days",
  "duration_median_days", "duration_p75_days", "duration_max_days",
  "duration_iqr_days", "suppression_applied"
)
distribution_required_columns <- c(
  "analysis_year", "stratification", "stratum_value", "utilization_category",
  "duration_days", "n_duration_events", "suppression_applied"
)

validate_columns <- function(data, data_name, columns) {
  missing_columns <- setdiff(columns, names(data))
  if (length(missing_columns) > 0L) {
    stop(
      data_name, " is missing required columns: ",
      paste(missing_columns, collapse = ", ")
    )
  }
}

coerce_numeric_columns <- function(data, columns) {
  for (column in intersect(columns, names(data))) {
    data[[column]] <- suppressWarnings(as.numeric(data[[column]]))
  }
  data
}

filter_to_analysis_year <- function(data, data_name) {
  data$analysis_year <- suppressWarnings(as.integer(data$analysis_year))
  filtered <- data[data$analysis_year == selected_analysis_year, , drop = FALSE]
  if (nrow(filtered) == 0L) {
    stop(data_name, " does not contain analysis_year = ", selected_analysis_year, ".")
  }
  filtered
}

category_order <- c(
  "Acute Inpatient",
  "Hospice (Inpatient)",
  "Skilled Nursing",
  "IPF",
  "Emergency Department"
)

stratification_order <- c(
  "overall",
  "mx_insurance_group",
  "mx_insurance_segment",
  "age_group",
  "sex",
  "cfi_level"
)

stratification_labels <- c(
  overall = "Overall Komodo",
  mx_insurance_group = "Insurance plan",
  mx_insurance_segment = "Insurance segment",
  age_group = "Age group",
  sex = "Sex",
  cfi_level = "CFI level"
)

utilization_palette <- c(
  "Acute Inpatient" = "#2E86AB",
  "Hospice (Inpatient)" = "#7D5BA6",
  "Skilled Nursing" = "#3F7C85",
  "IPF" = "#A44A3F",
  "Emergency Department" = "#F2B134"
)

ordered_stratum_values <- function(values, stratification) {
  values <- unique(as.character(values))
  values <- values[!is.na(values)]
  if (stratification == "overall") {
    return("Komodo")
  }
  if (stratification == "age_group") {
    preferred <- c("40-49", "50-64", "65-74", "75-84", "85+", "Unknown")
    return(c(intersect(preferred, values), setdiff(sort(values), preferred)))
  }
  if (stratification == "cfi_level") {
    preferred <- c("Robust", "Prefrail", "Frail", "Unknown")
    return(c(intersect(preferred, values), setdiff(sort(values), preferred)))
  }
  c(sort(setdiff(values, "Unknown")), if ("Unknown" %in% values) "Unknown")
}

prepare_summary <- function(data, data_name) {
  validate_columns(data, data_name, summary_required_columns)
  data <- filter_to_analysis_year(data, data_name)
  data <- coerce_numeric_columns(
    data,
    c(
      "analysis_year", "n_eligible_patient_years", "n_patients_with_visit",
      "pct_patient_years_with_visit", "n_visits", "duration_n",
      "duration_missing_n", "duration_missing_pct", "duration_mean_days",
      "duration_sd_days", "duration_min_days", "duration_p25_days",
      "duration_median_days", "duration_p75_days", "duration_max_days",
      "duration_iqr_days"
    )
  )
  data <- data[data$stratification %in% stratification_order, , drop = FALSE]
  data$utilization_category <- factor(
    data$utilization_category,
    levels = category_order
  )
  data
}

prepare_distribution <- function(data) {
  validate_columns(data, "duration distribution", distribution_required_columns)
  data <- filter_to_analysis_year(data, "duration distribution")
  data <- coerce_numeric_columns(data, c("duration_days", "n_duration_events"))
  data <- data[
    data$suppression_applied != "yes" &
      !is.na(data$duration_days) &
      !is.na(data$n_duration_events) &
      data$n_duration_events >= 11,
    ,
    drop = FALSE
  ]
  data$utilization_category <- factor(
    data$utilization_category,
    levels = category_order
  )
  data
}

overall_inputs <- prepare_summary(
  read_required_csv(
    paste0("6.4_healthcare_utilization_summary_", selected_analysis_year, ".csv")
  ),
  "overall utilization summary"
)
subgroup_inputs <- prepare_summary(
  read_required_csv(
    paste0(
      "6.4_healthcare_utilization_summary_by_subgroup_",
      selected_analysis_year,
      ".csv"
    )
  ),
  "subgroup utilization summary"
)
summary_inputs <- rbind(overall_inputs, subgroup_inputs)

duration_inputs <- prepare_distribution(
  read_required_csv(
    paste0(
      "6.4_healthcare_utilization_duration_distribution_",
      selected_analysis_year,
      ".csv"
    )
  )
)

observed_stratifications <- unique(summary_inputs$stratification)
missing_stratifications <- setdiff(stratification_order, observed_stratifications)
if (length(missing_stratifications) > 0L) {
  stop(
    "Summary inputs are missing requested stratifications: ",
    paste(missing_stratifications, collapse = ", ")
  )
}

format_value <- function(value, digits = 1) {
  ifelse(is.na(value), "Suppressed", formatC(value, format = "f", digits = digits))
}

format_n_pct <- function(n, pct) {
  ifelse(
    is.na(n) | is.na(pct),
    "Suppressed",
    paste0(formatC(n, format = "f", digits = 0, big.mark = ","),
           " (", formatC(pct, format = "f", digits = 1), "%)")
  )
}

format_mean_sd <- function(mean, sd) {
  ifelse(
    is.na(mean),
    "Suppressed",
    paste0(format_value(mean), " (", format_value(sd), ")")
  )
}

format_median_iqr <- function(median, iqr) {
  ifelse(
    is.na(median),
    "Suppressed",
    paste0(format_value(median), " (", format_value(iqr), ")")
  )
}

make_statistics_table <- function(data, stratification, utilization_category = NULL) {
  table_data <- data[data$stratification == stratification, , drop = FALSE]
  if (!is.null(utilization_category)) {
    table_data <- table_data[
      table_data$utilization_category == utilization_category,
      ,
      drop = FALSE
    ]
  }
  if (nrow(table_data) == 0L) {
    return(NULL)
  }
  table_data$stratum_value <- factor(
    table_data$stratum_value,
    levels = ordered_stratum_values(table_data$stratum_value, stratification)
  )
  table_data <- table_data[order(table_data$stratum_value, table_data$utilization_category), , drop = FALSE]
  output <- data.frame(
    Patients = format_n_pct(
      table_data$n_patients_with_visit,
      table_data$pct_patient_years_with_visit
    ),
    `Mean (SD) days` = format_mean_sd(
      table_data$duration_mean_days,
      table_data$duration_sd_days
    ),
    `Median (IQR) days` = format_median_iqr(
      table_data$duration_median_days,
      table_data$duration_iqr_days
    ),
    `Min days` = format_value(table_data$duration_min_days),
    `Max days` = format_value(table_data$duration_max_days),
    check.names = FALSE,
    stringsAsFactors = FALSE
  )
  if (stratification == "overall") {
    output <- cbind(
      `Visit type` = as.character(table_data$utilization_category),
      output
    )
  } else {
    output <- cbind(
      Subgroup = as.character(table_data$stratum_value),
      output
    )
  }
  knitr::kable(output, align = c("l", "r", "r", "r", "r", "r"))
}

plot_prevalence <- function(data, stratification, utilization_category = NULL) {
  plot_data <- data[data$stratification == stratification, , drop = FALSE]
  if (!is.null(utilization_category)) {
    plot_data <- plot_data[
      plot_data$utilization_category == utilization_category,
      ,
      drop = FALSE
    ]
  }
  plot_data <- plot_data[!is.na(plot_data$pct_patient_years_with_visit), , drop = FALSE]
  if (nrow(plot_data) == 0L) {
    return(NULL)
  }
  if (stratification == "overall") {
    plot_data$plot_x <- factor(
      plot_data$utilization_category,
      levels = category_order
    )
  } else {
    plot_data$plot_x <- factor(
      plot_data$stratum_value,
      levels = ordered_stratum_values(plot_data$stratum_value, stratification)
    )
  }
  plot_data$plot_label <- format_n_pct(
    plot_data$n_patients_with_visit,
    plot_data$pct_patient_years_with_visit
  )

  ggplot(
    plot_data,
    aes(x = plot_x, y = pct_patient_years_with_visit, fill = utilization_category)
  ) +
    geom_col(
      position = position_dodge(width = 0.82),
      width = 0.75,
      color = "white",
      linewidth = 0.15
    ) +
    geom_text(
      aes(label = plot_label),
      position = position_dodge(width = 0.82),
      vjust = -0.25,
      size = 3.5,
      angle = 45,
      hjust = 0
    ) +
    scale_y_continuous(
      labels = function(x) paste0(formatC(x, format = "f", digits = 1), "%"),
      expand = expansion(mult = c(0, 0.24))
    ) +
    scale_x_discrete(labels = function(x) {
      vapply(
        x,
        function(value) paste(strwrap(value, width = 14), collapse = "\n"),
        character(1)
      )
    }) +
    scale_fill_manual(values = utilization_palette, drop = FALSE) +
    labs(
      title = if (stratification == "overall") {
        "Overall visit prevalence in Komodo"
      } else {
        paste(
          utilization_category,
          "prevalence by",
          stratification_labels[[stratification]]
        )
      },
      x = if (stratification == "overall") "Visit type" else stratification_labels[[stratification]],
      y = "Eligible patient-years with at least one visit",
      fill = "Visit type"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 11, angle = 35, hjust = 1),
      axis.text.y = element_text(size = 11),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 15, face = "bold"),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold")
    ) +
    guides(fill = "none")
}

plot_duration_box <- function(summary_data, distribution_data, stratification, utilization_category = NULL) {
  box_data <- summary_data[summary_data$stratification == stratification, , drop = FALSE]
  point_data <- distribution_data[
    distribution_data$stratification == stratification,
    ,
    drop = FALSE
  ]
  if (!is.null(utilization_category)) {
    box_data <- box_data[
      box_data$utilization_category == utilization_category,
      ,
      drop = FALSE
    ]
    point_data <- point_data[
      point_data$utilization_category == utilization_category,
      ,
      drop = FALSE
    ]
  }
  box_data <- box_data[
    !is.na(box_data$duration_min_days) &
      !is.na(box_data$duration_p25_days) &
      !is.na(box_data$duration_median_days) &
      !is.na(box_data$duration_p75_days) &
      !is.na(box_data$duration_max_days),
    ,
    drop = FALSE
  ]
  if (nrow(box_data) == 0L) {
    return(NULL)
  }
  if (stratification == "overall") {
    box_data$plot_x <- factor(box_data$utilization_category, levels = category_order)
    point_data$plot_x <- factor(point_data$utilization_category, levels = category_order)
  } else {
    stratum_levels <- ordered_stratum_values(box_data$stratum_value, stratification)
    box_data$plot_x <- factor(box_data$stratum_value, levels = stratum_levels)
    point_data$plot_x <- factor(point_data$stratum_value, levels = stratum_levels)
  }
  box_data$median_label <- paste0("Median = ", format_value(box_data$duration_median_days))
  box_data$q1_label <- paste0("Q1 = ", format_value(box_data$duration_p25_days))
  box_data$q3_label <- paste0("Q3 = ", format_value(box_data$duration_p75_days))

  box_plot <- ggplot(
    box_data,
    aes(x = plot_x, color = utilization_category, fill = utilization_category)
  ) +
    geom_boxplot(
      aes(
        ymin = duration_min_days,
        lower = duration_p25_days,
        middle = duration_median_days,
        upper = duration_p75_days,
        ymax = duration_max_days
      ),
      stat = "identity",
      width = 0.62,
      alpha = 0.35,
      linewidth = 0.5
    ) +
    geom_text(
      data = box_data,
      aes(x = plot_x, y = duration_median_days, label = median_label),
      inherit.aes = FALSE,
      vjust = 1.8,
      size = 3.4,
      color = "black",
      fontface = "bold"
    ) +
    geom_text(
      data = box_data,
      aes(x = plot_x, y = duration_p25_days, label = q1_label),
      inherit.aes = FALSE,
      vjust = 1.8,
      size = 3.2,
      color = "black"
    ) +
    geom_text(
      data = box_data,
      aes(x = plot_x, y = duration_p75_days, label = q3_label),
      inherit.aes = FALSE,
      vjust = -0.7,
      size = 3.2,
      color = "black"
    ) +
    geom_point(
      data = point_data,
      aes(
        x = plot_x,
        y = duration_days,
        color = utilization_category,
        size = n_duration_events
      ),
      position = position_jitter(width = 0.12, height = 0),
      alpha = 0.55,
      inherit.aes = FALSE
    ) +
    scale_color_manual(values = utilization_palette, drop = FALSE) +
    scale_fill_manual(values = utilization_palette, drop = FALSE) +
    scale_size_area(max_size = 4, guide = "none") +
    scale_x_discrete(labels = function(x) {
      vapply(
        x,
        function(value) paste(strwrap(value, width = 14), collapse = "\n"),
        character(1)
      )
    }) +
    labs(
      title = if (stratification == "overall") {
        "Overall duration by visit type in Komodo"
      } else {
        paste(
          utilization_category,
          "duration by",
          stratification_labels[[stratification]]
        )
      },
      subtitle = "Boxplot shows min, quartiles, median, and max; points show aggregate duration frequencies",
      x = if (stratification == "overall") "Visit type" else stratification_labels[[stratification]],
      y = "Duration (days)",
      color = "Visit type"
    ) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(size = 11, angle = 35, hjust = 1),
      axis.text.y = element_text(size = 11),
      axis.title = element_text(size = 12),
      plot.title = element_text(size = 15, face = "bold"),
      plot.subtitle = element_text(size = 11),
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold")
    )
  box_plot
}

draw_side_by_side <- function(left_plot, right_plot, left_message, right_message) {
  grid::grid.newpage()
  grid::pushViewport(
    grid::viewport(
      layout = grid::grid.layout(
        nrow = 1,
        ncol = 2,
        widths = grid::unit(c(1, 1), "null")
      )
    )
  )
  for (plot_spec in list(
    list(plot = left_plot, message = left_message, column = 1),
    list(plot = right_plot, message = right_message, column = 2)
  )) {
    grid::pushViewport(
      grid::viewport(layout.pos.row = 1, layout.pos.col = plot_spec$column)
    )
    if (is.null(plot_spec$plot)) {
      grid::grid.text(plot_spec$message, gp = grid::gpar(fontsize = 12))
    } else {
      print(plot_spec$plot, newpage = FALSE)
    }
    grid::popViewport()
  }
  grid::popViewport()
  invisible(NULL)
}
```

# Overall Komodo

The percentage is the percentage of eligible patient-years with at least one
visit of the specified type. Duration summaries use nonmissing duration values
from prepared utilization events. Scatter points represent unsuppressed
aggregate duration frequencies and do not represent locally loaded
patient-level rows.

```{r overall-plots, results='asis', fig.show='asis', fig.width=16, fig.height=8}
draw_side_by_side(
  plot_prevalence(summary_inputs, "overall"),
  plot_duration_box(summary_inputs, duration_inputs, "overall"),
  "No unsuppressed overall visit-prevalence rows are available.",
  "No unsuppressed overall duration rows are available."
)
cat("\n\n### Summary statistics\n\n")
overall_statistics_table <- make_statistics_table(summary_inputs, "overall")
if (is.null(overall_statistics_table)) {
  cat("No overall summary rows are available.\n\n")
} else {
  print(overall_statistics_table)
  cat("\n\n")
}
```

# Subgroup Analyses

Each subgroup analysis is presented separately for each visit type. This keeps
the x-axis focused on the subgroup categories and gives each visit type its own
plot and statistics table.

```{r subgroup-plots, results='asis', fig.show='asis', fig.width=16, fig.height=8}
for (stratification in setdiff(stratification_order, "overall")) {
  cat("\n\n## ", stratification_labels[[stratification]], "\n\n", sep = "")

  for (utilization_category in category_order) {
    cat("\n\n### ", utilization_category, "\n\n", sep = "")

    draw_side_by_side(
      plot_prevalence(summary_inputs, stratification, utilization_category),
      plot_duration_box(
        summary_inputs,
        duration_inputs,
        stratification,
        utilization_category
      ),
      "No unsuppressed visit-prevalence rows are available for this visit type.",
      "No unsuppressed duration rows are available for this visit type."
    )

    cat("\n\n#### Summary statistics\n\n")
    statistics_table <- make_statistics_table(
      summary_inputs,
      stratification,
      utilization_category
    )
    if (is.null(statistics_table)) {
      cat("No summary rows are available for this visit type.\n\n")
    } else {
      print(statistics_table)
      cat("\n\n")
    }
  }
}
```
